---
title: "Showcase"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("./R/CoreModel.R", local = knitr::knit_global())
source("./R/Defaults.R", local = knitr::knit_global())
source("./R/DifferentialEquations.R", local = knitr::knit_global())
source("./R/SupportingFunctions.R", local = knitr::knit_global())
```

## Illustration 1: Simulating from the model

The simPanic function allows you to simulates the model's behavior for the specified amount of time (here, 60 "minutes" of simulated time). 

```{r, results='hide'}
out <- simPanic(time = 0:60, stepsize = .001)
```

The output file has three components:

  1. The initial values and parameters used in the simulation (i.e., input).
  2. Adherence to exposure exercises when a treatment is administered (i.e., adherence; see Illustration 4).
  3. A matrix where rows are time steps and columns are variables (i.e., outmat).


```{r, results='hide'}
out$input             
out$adherence 
out$outmat  
```

```{r}
dim(out$outmat)
```

For the 'outmat' output file we have a matrix of 61 time steps (0 through 60) for 12 variables, including:

* 5 Fast-changing Model Components: 
  + A = Arousal; 
  + PT = Perceived Threat; 
  + E = Escape; 
  + N = Noise; 
  + H = Homeostatic Feedback
* 3 Slow-changing Model Components: 
  + S = Arousal Schema; 
  + X = Escape Schema; 
  + V = Avoidance
* 1 Emergent Component: 
  + AF = Fear
* 2 model parameters

### Visualize Simulation Results for Key Model Components
We can visualize the simulation results for some of the key model components. 

```{r}
results <- out$outmat
plot.new()
plot.window(xlim = c(0, length(results$A)),  ylim=c(-.75, 1))
axis(1); axis(2,las=2)
mtext("Minutes", side = 1, line=3, cex=1.25)
lines(results$E, col="red",lwd=3)
lines(results$H, col="red",lty="dashed",lwd=3)
lines(results$A, col="grey",lwd=3)
lines(results$PT, col="black",lwd=3)
legend("topleft", legend=c("Arousal","Perceived Threat","Escape","Homeostatic Feedback"),
       col = c("grey","black","red","red"),
       lty = c(1,1,1,2), bty="n",cex=1.25)
```

## Illustration 2: Initial Values & Parameters
The initial values of model components and parameters defining the relationships among model components can be changed in order to evaluate how the model behaves under different conditions. If left unspecified, the simPanic will use default initial values and parameters. 

To specify initial values or parameters of interest, first create a list assigning the chosen value for the relevant component/parameter. Here, we specify an initial value of 1 for AS and we specify a value of .55 for the parameter r_A, which in the equation defines the rate of change for arousal A:

```{r}
initial_specified <- list("S"= 1)
parameters_specified <- list("A" = list("r_A"= .55))
```

Then use that list to define initial values (initial = ) and parameters (parameters = ) in the simPanic function: 

```{r, results='hide'}
out2 <- simPanic(time = 0:120,
                stepsize = .001,
                initial = initial_specified,          
                parameters = parameters_specified) 
```

To be sure, we check the input initial values and parameters:

```{r}
out2$input$initial$S
out2$input$parameters$A$r_A 
```

We can see that, as specified, the initial S = 1 and the parameter r_A = .55. 

### Visualize Simulation Results for Key Model Components
We can visualize the simulation results for some of the key model components. 

```{r}
results2 <- out2$outmat
plot.new()
plot.window(xlim = c(0, length(results2$A)),  ylim=c(-.75, 1))
axis(1); axis(2,las=2)
mtext("Minutes", side = 1, line = 3, cex = 1.25)
lines(results2$E, col = "red", lwd = 3)
lines(results2$H, col = "red",lty = "dashed", lwd = 3)
lines(results2$A, col = "grey",lwd = 3)
lines(results2$PT, col = "black",lwd = 3)
legend("bottomleft", legend = c("Arousal","Perceived Threat","Escape","Homeostatic Feedback"),
       col = c("grey", "black", "red", "red"),
       lty = c(1,1,1,2), bty = "n",cex = 1.25)
```

## Illustration 3: Building in a Perturbation
A perturbation to Arousal can be added to the model to evaluate how the system responds to a perturbation of given strength at a given time. 

To do so, we specify that the perturbation should occur at minute 10 and the perturbation should move arousal to A=0.60. 

```{r}
perturb_time <- 10
perturb_strength <- .60

parameters_specified <- list("Tx" = list("minuteP" = perturb_time, "strengthP" = perturb_strength))
initial_specified <- list("S" = .60)
```

Then we use that list to define initial values (initial = ) and parameters (parameters = ) in the simPanic function: 

```{r, results='hide'}
out3  <- simPanic(time = 0:60, # 1h
                  stepsize = .001,
                  parameters = parameters_specified,
                  initial = initial_specified,
                  pbar = TRUE)
```

### Visualize Simulation Results for Key Model Components
```{r}
results3 <- out3$outmat
plot.new()
plot.window(xlim = c(0, length(results3$A)),  ylim = c(-.75, 1))
axis(1); axis(2, las = 2)
mtext("Minutes", side = 1, line = 3,cex = 1.25)
rect(perturb_time, -.5, (perturb_time+1), perturb_strength, col = "grey90", border = FALSE)
text(perturb_time, -0.6, "Perturbation")
lines(results3$E, col = "red", lwd = 3)
lines(results3$H, col = "red", lty = "dashed", lwd = 3)
lines(results3$A, col = "grey", lwd = 3)
lines(results3$PT, col = "black", lwd = 3)
legend("topright", legend = c("Arousal", "Perceived Threat", "Escape", "Homeostatic Feedback"),
       col = c("grey", "black", "red", "red"),
       lty = c(1,1,1,2), bty = "n")
```

Note that the results of these perturbations are affected by stochastic variation in arousal. As a result, the response to the perturbation can be different each time the simulation is run. To illustrate, we can repeat the simulation 10 times and plot the trajectory of arousal for each.

```{r}
N_sim <- 10

# Storage for the results:
results <- list()  

for(i in 1:N_sim) {
  out <- simPanic(time = 0:50,
                  stepsize = .001,
                  parameters = parameters_specified,
                  initial = initial_specified,
                  pbar = FALSE)

  results[[i]] <- out
}

plot.new()
plot.window(xlim = c(0, length(results[[1]]$outmat$AF)),  ylim=c(-.75, 1))
axis(1); axis(2,las=2)
mtext("Minutes", side = 1, line = 3,cex = 1.25)
rect(perturb_time, -.5, (perturb_time+1), perturb_strength, col = "grey90", border = FALSE)
text(perturb_time,-0.6,"Perturbation")
for(i in 1:length(results)){lines(results[[i]]$outmat$AF, col = "black", lwd = .5)}
legend("topright", legend = c("Fear"),
       col = c("black"),
       lty = c(1), bty="n")
```

## Illustration 4: Iterating Across Parameters
By iterating across different initial values and parameters, we can evaluate how the system behaves under different conditions.

For example, we can examine how the system will respond to a range of perturbation strengths at different values of arousal schema (S). 

First, we specify the number of simulations you want to run per condition:
```{r}
N_sim <- 10
```

Second, we generate a random set of perturbation strengths between 0 and 1 and specify the Arousal Schema (S) values you want to examine: 
```{r}
perturb_strengths <- runif(N_sim, 0, 1)  
Ss <- c(.10, .50)  
```

Third, we wet the time of the perturbation and the total amount of simulated time:
```{r}
perturb_time <- 10 # 10 minutes
time <- 0:60 # 1h 
```

Finally we can run and store the simulations: 
```{r}
results <- list()

for (h in 1:length(Ss)){
  initial_specified <- list("S" = Ss[h])

  for (i in 1:N_sim) {
    perturb_strength <- perturb_strengths[i]
    parameters_specified <- list("Tx" = list("minuteP"= perturb_time, "strengthP" = perturb_strength))

    out <- simPanic(time = time,
                    stepsize = .001,
                    parameters = parameters_specified,
                    initial = initial_specified,
                    pbar = FALSE)

    results[[((h-1)*N_sim)+i]] <- out
  }
}
```

From the results we compute the maximum fear (AF) in response to the perturbations in the two conditions:
```{r}
AFs_1 <- AFs_2 <- vector()
for (i in 1:N_sim){
  AFs_1[i] <- max(results[[i]]$outmat$AF)
  AFs_2[i] <- max(results[[((length(Ss)-1)*N_sim)+i]]$outmat$AF)
}
```

### Visualize Simulation Results for Key Model Components
```{r}
plot.new()
plot.window(xlim=c(0,1),ylim=c(0,1))
axis(1); axis(2,las=2)
mtext("Perturb Strength", side = 1, line = 3, cex = 1.25)
mtext("Peak Fear", side = 2, line = 3, srt = 0, cex = 1.25)
points(perturb_strengths, AFs_1, pch = 19, col = rgb(.3, .3, .3, alpha = .3))
points(perturb_strengths, AFs_2, pch = 19, col = rgb(0, 0, 1, alpha = .3))
legend("topleft", 1, legend = c(paste("S =", Ss[1]), paste("S =", Ss[2])),
       col = c("grey", "blue"),
       pch = 19, bty = "n")
```

## Illustration 5: Building in Treatment
Five Cognitive Behavioral Therapy (CBT) Intervention Components are built in to the model:

1. I1 = Psychoeducation targeting Arousal Schema (S)
2. I2 = Psychoeducation targeting Escape Schema (X)
3. I3 = Cognitive Restructuring targeting Arousal Schema (S)
4. I4 = Interoceptive Exposure
5. I5 = In Vivo Exposure

To incorporate the intervention components into the simulation, specify a baseline period (in days and in weeks) and then specify when the treatment components should be delivered. Here we specify a baseline period of 0 weeks and 1 day: 

```{r}
baseline_weeks <- 0
baseline_days <- 1
```

Furthermore we specify that psychoeducation targeting S (I1) should occurs at Day 2; psychoeducation targeting X (I2) should occurs at Day 2; cognitive restructing targeting S (I3) should occurs at Day 9 and 16; interoceptive exposure (I4) should occur at Days 9-36; and in vivo exposure (I5) should occur at Days 23-36. 

```{r}
tx <- list("I1" = baseline_days+baseline_weeks*7+c(1),
           "I2" = baseline_days+baseline_weeks*7+c(1), 
           "I3" = (baseline_days+baseline_weeks+(1:2)*7)+c(1), 
           "I4" = (baseline_days+baseline_weeks*7)+7+(1:(7*4)),
           "I5" = baseline_days+(baseline_weeks+3)*7+(1:14)[rep(c(TRUE,TRUE), 7)]) 
```

To incorporate variability in treatment efficacy and adherence, you can specify three parameters: 

* I123_alpha determines the efficacy (in % reduction) for I1-3
* I4Adh determines the adherence (in % likelihood of completing) for I4-I5
* I4RdEs determines the Extent to which instruction to refrain from escape behavior raises threshold at which perceived threat will lead to escape (0-1)

```{r}
parameters_specified <- list("Tx" = list("I123_alpha" = rbeta(n = 1, 1, 9), 
                                         "I4Adh" = rbeta(n = 1, 2, 2/3),
                                         "I4RdEs" = rbeta(n = 1, 2, 2/3))) 
```

Ensure there is enough time for all treatment components to be delivered: 
```{r}
time <- 0:(60*24*7*5)
```

Finally, given that this is a treatment, specify S and X values consistent with panic disorder: 
```{r}
initial_specified <- list("S" = .80, "X" = .20)
```

Use the simPanic function to simulate the specified treatment:
```{r, results='hide'}
out4 <- simPanic(time = time,
                  stepsize = .001,
                  parameters = parameters_specified,
                  initial = initial_specified,
                  tx = tx,
                  pbar = TRUE)
```

Adherence gives a vector of all assigned exposure exercises, where 1=Completed. 0=Not Completed. The mean of this vector gives the proportion of completed exposures. 

```{r}
out4$adherence
mean(out4$adherence)
```

### Visualize Simulation Results for Key Model Components

```{r}
results4 <- out4$outmat
endbase <- (60*24*7) * baseline_weeks
endtreat <- endbase + (60*24*7)*5
exposures_x <- out4$adherence*(endbase+(7*60*24) + seq(1:length(out4$adherence))*60*24)
exposures_x[exposures_x==0] <- NA
exposures_y <- rep(0, length(exposures_x))
cog_x <- c(endbase + (7*60*24), endbase + (14*60*24))
cog_y<-rep(-.025, length(cog_x))
psyed_x <- c(endbase + (0*60*24), endbase + (0*60*24))
psyed_y <- rep(-.025, length(psyed_x))
cols <- vector()
cols[1] <- rgb(1.0, 0, 0, alpha = .05)
cols[2] <- rgb(0, 0, 1.0, alpha = .05)

plot.new()
plot.window(xlim = c(0, length(results4$A)),  ylim = c(0, 1))
axis(1, at = seq(endbase, length(results4$A), by = (60*24*7)), labels = c("S1", "S2", "S3", "S4", "S5", "E"))
axis(2, at = seq(0, 1, .25), las = 2)
mtext("Weeks", side = 1, line = 4, cex = 1.25)
polygon(c(seq(1, length(results4$AS)), rev(seq(1,length(results4$S)))), c(results4$S, rep(0,length(results4$S))), col = cols[1], border = NA)
polygon(c(seq(1, length(results4$ES)), rev(seq(1,length(results4$X)))), c(results4$X, rep(0,length(results4$X))), col = cols[2], border = NA)
lines(results4$PT, col = "dark grey")
lines(results4$S, col = "red", lwd = 3)
lines(results4$X, col = "blue", lwd = 3)
points(exposures_x, exposures_y, pch = 19, col = "black")
points(cog_x, cog_y, pch = 15, col = "black")
points(psyed_x, psyed_y, pch = 17, col = "black")
legend("topright", 1, legend = c("Arousal Schema","Escape Schema"),
       col = c("red", "blue"),
       lty = 1, bty = "n")
legend("topleft", 1, legend = c("Psychoeducation", "Cognitive Restructuring", "Exposure Exercise"),
       pch = c(17, 15, 19), bty = "n")
```